Verification of the CVM algorithm with a New Recursive Analysis Technique

### 2025-01-25 isabelle/CVM_Subsample/document/appendix.tex
Also verify that state_chi allways contains below thresh elements.  + Fix minor typos.

\section{Informal Proof}
This section includes an informal version of the proof for the tail bounds and unbiasedness
of the abstract algorithm (Algorithm~\ref{alg:cvm_abs}) for interested readers.

@@

\begin{lemma}
\label{le:neg_cor_prelim}
Let $\phi : (0,1] \times \{0,1\} \rightarrow \mathbb R_{\geq 0}$ be a function, fulfilling the following
  conditions:
\begin{enumerate}
\item \label{cond:phi_1} $(1-\alpha) \phi(x,0) + \alpha \phi(x,1) \leq \phi(x/\alpha,1)$ for all
  $0 < \alpha < 1$, $0 < x \leq 1$, and
\item \label{cond:phi_2} $\phi(x,0) \leq \phi(y,0)$ for all $0 < x < y \leq 1$.
\end{enumerate}
Then for all 
$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$, $\Omega \in \{\Omega_k,\Omega_k'\}$:
\[
  \expect_{\Omega}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] \leq \phi(1,1)^{|S|} 
\]
\end{lemma}
\begin{proof}
We show the result using induction over $k$. Note that we show the statement for arbitrary $S$,
i.e., the induction statements are:
\begin{eqnarray*}
P(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \;
  \expect_{\Omega_k}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right]
    \leq \phi(1,1)^{|S|} \right) \\
Q(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; 
  \expect_{\Omega'_k}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right]
    \leq \phi(1,1)^{|S|} \right)
\end{eqnarray*}
and we will show $P(0),Q(1),P(1),Q(2),P(2),\ldots,Q(l),P(l)$ successively.
\paragraph{Induction start $P(0)$:} \phantom{.}\\
We have $S \subseteq \emptyset$, and hence
\[
\expect_{\Omega_0}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] =
\expect_{\Omega_0}\left[1\right] = 1 \leq \phi(1,1)^0 \textrm{.}
\]
\paragraph{Induction step $P(k) \rightarrow Q(k+1)$:} \phantom{.}\\
Let $S \subseteq \{ a_1, \ldots, a_{k+1} \}$ and define $S' := S - \{ a_{k+1} \}$.
Note that $\Omega'_{k+1}$ can be constructed from $\Omega_k$ as a compound distribution, where
$a_{k+1}$ is included in the buffer, with the probability $p$, which is itself a random variable
over the space $\Omega_k$.

@@

A corollary and more practical version of the previous lemma is:
\begin{lemma}\label{le:neg_cor_neg}
Let $q \leq 1$ and $h : \mathbb [0,q^{-1}] \rightarrow \mathbb R_{\geq 0}$ be concave then for all 
$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$, $\Omega \in \{\Omega_k,\Omega_k'\}$:
\[
  \expect_{\Omega}\left[ \prod_{s \in S} I(p>q) h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} 
\]
\end{lemma}
\begin{proof}
Follows from Lemma~\ref{le:neg_cor_prelim} for $\phi(p,\tau) := I(p > q) h(\tau p^{-1})$. We just
need to check Conditions 1 and 2. Indeed,
\begin{eqnarray*}
  (1-\alpha) \phi(x,0) + \alpha \phi(x,1) & = & (1-\alpha) I(x>q) h(0) + \alpha I(x>q) h(x^{-1}) \\
   & \leq & I(x > q) h(\alpha x^{-1}) \leq I(x > q \alpha) h(\alpha x^{-1}) = \phi(x/\alpha,1)
\end{eqnarray*}
for $0 < \alpha < 1$ and $0< x \leq 1$, where we used that $x > q$ implies $x > q \alpha$; and
\[
  \phi(x,0) = I(x>q) h(0) \leq I(y > q) h(0) = \phi(y,0)
\]
for $0 < x < y \leq 1$, where we used that $x > q$ implies $y > q$.
\end{proof}

@@

\subsection{Concentration}
Let us now see how we can obtain concentration bounds using Lemma~\ref{le:neg_cor_neg}, i.e., that
the result of the algorithm is concentrated around the cardinality of $A = \{ a_1, \ldots, a_l \}$.
This will be done using the Cram\'{e}r--Chernoff method for the probability that the estimate is 
above $(1+\eps)|A|$ (resp.~below $(1-\eps)|A|$) assuming $p$ is not too small and a tail estimate 
for $p$ being too small.

@@

To get a tail estimate, we use the Cram\'{e}r--Chernoff method:
{\allowdisplaybreaks
\begin{eqnarray*}
  L & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq
    \exp(t (1+\eps) |A|) \wedge p \geq q \right) \\
  & \leq &
    \prob_{\Omega} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \right) \\
  & \underset{\textrm{Markov}}{\leq} &
    \exp( -t (1+\eps) |A| ) \expect_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
 & \leq & \exp( -t (1+\eps) |A| ) 
  \expect_\Omega \left[ \prod_{s \in A'} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
 & \leq & \exp( -t (1+\eps) |A| ) 
  \expect_\Omega \left[ \prod_{s \in A'} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
 & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
 & \underset{h(1)\geq 1}{\leq} & \left(\exp( \ln(h(1)) -t (1+\eps))\right) ^ {|A|} 
\end{eqnarray*}}
So we just need to show that (using $|A|=\frac{n}{4q}$):
\[
  \ln(h(1))-t (1+\eps) \leq \frac{-q \eps^2}{3}
\]
The latter can be established by analyzing the function
\[
  f(\eps) := -\ln(1+q \eps) + q \ln(1+\eps) (1+\eps) - \frac{q \eps^2}{3} =
    -\ln (h(1)) +t (1+\eps) - \frac{q\eps^2}{3} \textrm{.}
\]
For which it is easy to check $f(0) = 0$ and the derivative with respect to $\eps$ is non-negative
in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e., $f(\eps) \geq 0$.
\end{proof}

@@

With these definitions we again follow the Cram\'{e}r--Chernoff method:
{\allowdisplaybreaks
\begin{eqnarray*}
  L & \underset{t<0}{=} & \prob_{\Omega_l} \left( \exp(t p^{-1} |\chi|) \geq
    \exp(t (1-\eps)|A|) \wedge p \geq q \right) \\
    & \leq & \prob_{\Omega_l} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq
    \exp(t (1-\eps)|A|) \wedge p > q \right) \\
    & \underset{\textrm{Markov}}{\leq} & \exp( -t (1-\eps) |A| )
    \expect_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
    & = & \exp( -t (1-\eps) |A| ) \expect_\Omega \left[ \prod_{s \in A} I(p \geq q) 
      \exp ( t p^{-1} I(s \in \chi) ) \right] \\
    & \leq & \exp( -t (1-\eps) |A| ) \expect_\Omega 
      \left[ \prod_{s \in A} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
    & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1-\eps) |A| ) (h(1))^{|A|} \\
    & = & \exp( \ln(h(1)) -t (1-\eps) )^{|A|}
\end{eqnarray*}
}
Substituting $t$ and $h$ and using $|A| = \frac{n}{4q}$, we can see that the lemma is true if
\[
  f(\eps) := q \ln(1-\eps) (1-\eps) -\ln(1 - q \eps) - \frac{q}{2}\eps^2 =
  t (1-\eps) - \ln(h(1)) - \frac{q}{2} \eps^2
\]
is non-negative for $0 \leq q \leq \frac{1}{4}$ and $0 < \eps < 1$.
This can be verified by checking that $f(0) = 0$ and that the derivative with respect to $\eps$ is
non-negative.
\end{proof}
We can now establish the concentration result:
\begin{theorem}\label{th:concentration}
Let $0 < \eps < 1$ and $0 < \delta < 1$ and
$n \geq \frac{12}{\eps^2} \ln\left(\frac{3l}{\delta}\right)$ then:
\[
  L = \prob_{\Omega_l} \left( | p^{-1} |\chi| - |A| | \geq \eps |A| \right) \leq \delta
\]
\end{theorem}
\begin{proof}
Note that the theorem is trivial if $|A| < n$. If not:
\begin{eqnarray*}
  L & \leq & \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1-\eps) |A| \wedge p \geq q \right)\\
  & + &
    \prob_{\Omega_l} \left( | p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q \right) +
    \prob_{\Omega_l} \left( p < q \right) \\
    & \underset{Le.~\ref{le:upper_tail}-\ref{le:lower_tail}}{\leq} &
    \exp\left( - \frac{n}{8} \eps^2 \right)  + \exp\left( - \frac{n}{12} \eps^2 \right) + 
    l \exp\left( - \frac{n}{12} \right) \\
    & \leq & \frac{\delta}{3} + \frac{\delta}{3} + \frac{\delta}{3}
\end{eqnarray*}
\end{proof}
\subsection{Unbiasedness\label{sec:unbiasedness}}
Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g., we can choose $M = f^{-l}$).
Then we can derive from Lemma~\ref{le:neg_cor_neg} using $h(x) = x$ and $h(x) = M+1-x$ that
for all $s \in A$:
\begin{eqnarray*}
  \expect_{\Omega_l} [ p^{-1} I(s \in \chi) ] & = &
    \expect_{\Omega_l} [ I(p \geq M^{-1}) p^{-1} I(s \in \chi) ] \leq 1 \\
  \expect_{\Omega_l} [ M + 1 - p^{-1} I(s \in \chi) ] & = &
    \expect_{\Omega_l} [ I(p \geq M^{-1}) (M + 1 - p^{-1} I(s \in \chi)) ] \leq M
\end{eqnarray*}
which implies $\expect_{\Omega_l} [ p^{-1} I(s \in \chi) ] = 1$.
By linearity of expectation we conclude
\[
  \expect_{\Omega_l} [ p^{-1} |\chi| ] =
    \sum_{s \in A} \expect_{\Omega_l} [ p^{-1} I(s \in \chi) ] = |A| \textrm{.}
\]

>>>>

\section{Informal Proof\label{apx:informal_proof}}
This section includes an informal version of the proof for the tail bounds and unbiasedness
of the abstract algorithm (Algorithm~\ref{alg:cvm_abs}) for interested readers.

@@

\begin{lemma}
\label{le:prob_invariant}
Let $\phi : (0,1] \times \{0,1\} \rightarrow \mathbb R_{\geq 0}$ be a function, fulfilling the following
  conditions:
\begin{enumerate}
\item \label{cond:phi_1} $(1-\alpha) \phi(x,0) + \alpha \phi(x,1) \leq \phi(x/\alpha,1)$ for all
  $0 < \alpha < 1$, $0 < x \leq 1$, and
\item \label{cond:phi_2} $\phi(x,0) \leq \phi(y,0)$ for all $0 < x < y \leq 1$.
\end{enumerate}
Then for all 
$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$, $\Omega \in \{\Omega_k,\Omega_k'\}$:
\[
  \expect_{\Omega}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] \leq \phi(1,1)^{|S|} 
\]
\end{lemma}
\begin{proof}
We show the result using induction over $k$. Note that we show the statement for arbitrary $S$,
i.e., the induction statements are:
\begin{eqnarray*}
P(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \;
  \expect_{\Omega_k}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right]
    \leq \phi(1,1)^{|S|} \right) \\
Q(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; 
  \expect_{\Omega'_k}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right]
    \leq \phi(1,1)^{|S|} \right)
\end{eqnarray*}
and we will show $P(0),Q(1),P(1),Q(2),P(2),\ldots,Q(l),P(l)$ successively.
\paragraph{Induction start $P(0)$:} \phantom{.}\\
We have $S \subseteq \emptyset$, and hence
\[
\expect_{\Omega_0}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] =
\expect_{\Omega_0}\left[1\right] = 1 \leq \phi(1,1)^0 \textrm{.}
\]
\paragraph{Induction step $P(k) \rightarrow Q(k+1)$:} \phantom{.}\\
Let $S \subseteq \{ a_1, \ldots, a_{k+1} \}$ and define $S' := S - \{ a_{k+1} \}$.
Note that $\Omega'_{k+1}$ can be constructed from $\Omega_k$ as a compound distribution, where
$a_{k+1}$ is included in the buffer, with the probability $p$, which is itself a random variable
over the space $\Omega_k$.

@@

A corollary and more practical version of the previous lemma is:
\begin{lemma}\label{le:prob_invariant_simple}
Let $q \leq 1$ and $h : \mathbb [0,q^{-1}] \rightarrow \mathbb R_{\geq 0}$ be concave then for all 
$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$, $\Omega \in \{\Omega_k,\Omega_k'\}$:
\[
  \expect_{\Omega}\left[ \prod_{s \in S} I(p>q) h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} 
\]
\end{lemma}
\begin{proof}
Follows from Lemma~\ref{le:prob_invariant} for $\phi(p,\tau) := I(p > q) h(\tau p^{-1})$. We just
need to check Conditions 1 and 2. Indeed,
\begin{eqnarray*}
  (1-\alpha) \phi(x,0) + \alpha \phi(x,1) & = & (1-\alpha) I(x>q) h(0) + \alpha I(x>q) h(x^{-1}) \\
   & \leq & I(x > q) h(\alpha x^{-1}) \leq I(x > q \alpha) h(\alpha x^{-1}) = \phi(x/\alpha,1)
\end{eqnarray*}
for $0 < \alpha < 1$ and $0< x \leq 1$, where we used that $x > q$ implies $x > q \alpha$; and
\[
  \phi(x,0) = I(x>q) h(0) \leq I(y > q) h(0) = \phi(y,0)
\]
for $0 < x < y \leq 1$, where we used that $x > q$ implies $y > q$.
\end{proof}

@@

\subsection{Concentration}
Let us now see how we can obtain concentration bounds using Lemma~\ref{le:prob_invariant_simple}, i.e., that
the result of the algorithm is concentrated around the cardinality of $A = \{ a_1, \ldots, a_l \}$.
This will be done using the Cram\'{e}r--Chernoff method for the probability that the estimate is 
above $(1+\eps)|A|$ (resp.~below $(1-\eps)|A|$) assuming $p$ is not too small and a tail estimate 
for $p$ being too small.

@@

To get a tail estimate, we use the Cram\'{e}r--Chernoff method:
{\allowdisplaybreaks
\begin{eqnarray*}
  L & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq
    \exp(t (1+\eps) |A|) \wedge p \geq q \right) \\
  & \leq &
    \prob_{\Omega} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \right) \\
  & \underset{\textrm{Markov}}{\leq} &
    \exp( -t (1+\eps) |A| ) \expect_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
 & \leq & \exp( -t (1+\eps) |A| ) 
  \expect_\Omega \left[ \prod_{s \in A'} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
 & \leq & \exp( -t (1+\eps) |A| ) 
  \expect_\Omega \left[ \prod_{s \in A'} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
 & \underset{\textrm{Le.~\ref{le:prob_invariant_simple}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
 & \underset{h(1)\geq 1}{\leq} & \left(\exp( \ln(h(1)) -t (1+\eps))\right) ^ {|A|} 
\end{eqnarray*}}
So we just need to show that (using $|A|=\frac{n}{4q}$):
\[
  \ln(h(1))-t (1+\eps) \leq \frac{-q \eps^2}{3}
\]
The latter can be established by analyzing the function
\[
  f(\eps) := -\ln(1+q \eps) + q \ln(1+\eps) (1+\eps) - \frac{q \eps^2}{3} =
    -\ln (h(1)) +t (1+\eps) - \frac{q\eps^2}{3} \textrm{.}
\]
For which it is easy to check $f(0) = 0$ and the derivative with respect to $\eps$ is non-negative
in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e., $f(\eps) \geq 0$.
\end{proof}

@@

With these definitions we again follow the Cram\'{e}r--Chernoff method:
{\allowdisplaybreaks
\begin{eqnarray*}
  L & \underset{t<0}{=} & \prob_{\Omega_l} \left( \exp(t p^{-1} |\chi|) \geq
    \exp(t (1-\eps)|A|) \wedge p \geq q \right) \\
    & \leq & \prob_{\Omega_l} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq
    \exp(t (1-\eps)|A|) \wedge p > q \right) \\
    & \underset{\textrm{Markov}}{\leq} & \exp( -t (1-\eps) |A| )
    \expect_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
    & = & \exp( -t (1-\eps) |A| ) \expect_\Omega \left[ \prod_{s \in A} I(p \geq q) 
      \exp ( t p^{-1} I(s \in \chi) ) \right] \\
    & \leq & \exp( -t (1-\eps) |A| ) \expect_\Omega 
      \left[ \prod_{s \in A} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
    & \underset{\textrm{Le.~\ref{le:prob_invariant_simple}}}{\leq} & \exp( -t (1-\eps) |A| ) (h(1))^{|A|} \\
    & = & \exp( \ln(h(1)) -t (1-\eps) )^{|A|}
\end{eqnarray*}
}
Substituting $t$ and $h$ and using $|A| = \frac{n}{4q}$, we can see that the lemma is true if
\[
  f(\eps) := q \ln(1-\eps) (1-\eps) -\ln(1 - q \eps) - \frac{q}{2}\eps^2 =
  t (1-\eps) - \ln(h(1)) - \frac{q}{2} \eps^2
\]
is non-negative for $0 \leq q \leq \frac{1}{4}$ and $0 < \eps < 1$.
This can be verified by checking that $f(0) = 0$ and that the derivative with respect to $\eps$ is
non-negative.
\end{proof}
We can now establish the concentration result:
\begin{theorem}\label{th:concentration}
Let $0 < \eps < 1$ and $0 < \delta < 1$ and
$n \geq \frac{12}{\eps^2} \ln\left(\frac{3l}{\delta}\right)$ then:
\[
  L = \prob_{\Omega_l} \left( | p^{-1} |\chi| - |A| | \geq \eps |A| \right) \leq \delta
\]
\end{theorem}
\begin{proof}
Note that the theorem is trivial if $|A| < n$. If not:
\begin{eqnarray*}
  L & \leq & \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1-\eps) |A| \wedge p \geq q \right)\\
  & + &
    \prob_{\Omega_l} \left( | p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q \right) +
    \prob_{\Omega_l} \left( p < q \right) \\
    & \underset{Le.~\ref{le:upper_tail}-\ref{le:lower_tail}}{\leq} &
    \exp\left( - \frac{n}{8} \eps^2 \right)  + \exp\left( - \frac{n}{12} \eps^2 \right) + 
    l \exp\left( - \frac{n}{12} \right) \\
    & \leq & \frac{\delta}{3} + \frac{\delta}{3} + \frac{\delta}{3}
\end{eqnarray*}
\end{proof}
\subsection{Unbiasedness\label{sec:unbiasedness}}
Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g., we can choose $M = f^{-l}$).
Then we can derive from Lemma~\ref{le:prob_invariant_simple} using $h(x) = x$ and $h(x) = M+1-x$ that
for all $s \in A$:
\begin{eqnarray*}
  \expect_{\Omega_l} [ p^{-1} I(s \in \chi) ] & = &
    \expect_{\Omega_l} [ I(p \geq M^{-1}) p^{-1} I(s \in \chi) ] \leq 1 \\
  \expect_{\Omega_l} [ M + 1 - p^{-1} I(s \in \chi) ] & = &
    \expect_{\Omega_l} [ I(p \geq M^{-1}) (M + 1 - p^{-1} I(s \in \chi)) ] \leq M
\end{eqnarray*}
which implies $\expect_{\Omega_l} [ p^{-1} I(s \in \chi) ] = 1$.
By linearity of expectation we conclude
\[
  \expect_{\Omega_l} [ p^{-1} |\chi| ] =
    \sum_{s \in A} \expect_{\Omega_l} [ p^{-1} I(s \in \chi) ] = |A| \textrm{.}
\]
